Comparing summed probability distributions of shoreline and radiocarbon dates from the Mesolithic Skagerrak coast of Norway

### 2023-06-22 analysis/paper/paper.Rmd
Draft to Steinar

---
title: "Comparing summed probability distributions of shoreline and radiocarbon dates from the Mesolithic Skagerrak coast of Norway"
# author: 
# - 'Isak Roalkvam'
# - 'University of Oslo, Institute of Archaeology, Conservation and History'
  # - IsakRoalkvam:
  #     email: isak.roalkvam@iakh.uio.no
  #     institute: [iakh]
      # correspondence: true
#   - name: Steinar Solheim
#     email: steinar.solheim@khm.uio.no
#     institute: [mch]
#     correspondence: false
institute:
  - iakh: Institute of Archaeology, Conservation and History, University of Oslo
#   - mch: Museum of Cultural History, University of Oslo
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::pdf_document2:
      fig_caption: yes
      toc: false
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      # - --lua-filter=../templates/scholarly-metadata.lua
      # - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
header-includes:
  \usepackage{placeins}
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
# abstract: |
#   Text of abstract
# keywords: |
#   keyword 1; keyword 2; keyword 3
# highlights: |
#   These are the highlights. 
---

@@

knitr::opts_knit$set(
  eval.after = "fig.cap"
)

@@

# Introduction
Population size is regarded as an important driver of cultural variation, and is of critical importance to our understanding of past human societies [e.g. @french2021; @shennan2000]. The frequency distribution of radiocarbon dates has been used extensively as one proxy for past relative population sizes [e.g. @crema2022; @french2021], including in Norwegian Stone Age archaeology [@solheim2018; @solheim2020a; @nielsen2019; @nielsen2021; @bergsvik2021; @jorgensen2020; @lundstrom2021; see also @persson2009]. The potentially immense value of insights into past population dynamics, combined with the ubiquity of radiocarbon dates and the relative ease with which these can now be processed within what has been termed a dates as data methodology has undoubtedly contributed to the popularity of the approach. Several limitations and forms of criticism have, however, been directed at these procedures. Some of the objections are of a methodological nature, while others pertain to the underlying logic and the degree to which there is likely to be a direct connection between the frequency of ^14^C-dates and population dynamics [@carleton2021; @torfing2015]. What appears to be agreed upon by practitioners and critics alike is that radiocarbon dates are best analysed in this manner when compared and contrasted to other proxies for past population dynamics and other variables that might impact these [@french2016; @palmisano2017; @palmisano2021; @bergsvik2021]. This paper reports on and begins to unpack the relationship between two measures that have been linked to relative population size in the context of the Mesolithic Skagerrak coast in south-eastern Norway, namely summed probability distribution of calibrated radiocarbon dates and the summed probability distribution of shoreline dated sites (hereafter RSPD and SSPD, respectively).

@@

The final RSPD was subjected to the standard null-hypothesis testing approach through Monte Carlo simulation, as introduced above [see @shennan2013; @timpson2014], by fitting an exponential, logistic and uniform model to the observed RSPD. Following @timpson2021, the model parameters were identified by MLE search, using the differential evolution optimization algorithm *DEoptimR*. The Monte Carlo simulations were then performed using the *SPDsimulationTest()* function from *ADMUR*. For each model, a series of individual calendar years are drawn from the model distribution, with replacement, the number of which equals the number of bins in the observed RSPD. These are then 'uncalibrated' to a single value on the ^14^C scale and a random error from among the observed errors is added to the date. These are then calibrated back to the calendar scale and finally summed. Here, this procedure was repeated 10000 times for each null model. The 2.5th and 97.5th quantile of the resulting summed probability for each year across all simulations are then retrieved to create the 95% critical envelope with which to compare the observed RSPD. The degree to which the observed RSPD deviates from the critical envelope is then used to calculate a global p-value, indicating whether or not the null can be rejected.

@@

Summing the probability of the shoreline dated sites and the model-fitting procedures followed the same structure as that for radiocarbon dates and was partly based on re-purposed programming code from *rcarbon*. However, idiosyncrasies in the dating method did necessitate some adjustments. To illustrate this, the procedure for shoreline dating a single site, as suggested in @roalkvam2023, is provided in Figure \@ref(fig:shoredate) and outlined below.

@@

With an estimated trajectory of shoreline displacement at the site to be dated, the elevation of the site can be drawn on to find a likely date for when it was in use, under the assumption that it was in use when located on or close to the shoreline. Figure \@ref(fig:shoredate)C indicates where the elevation of the example site intersects the interpolated displacement curve. As it can be reasonably be assumed that a site was not in use when located under water, this effectively represents a *terminus post quem* (TPQ) date. However, a fairly common approach to shoreline dating has been find where the elevation of the site intersects the mean of the upper and lower limit of the displacement curve, and then adding a somewhat arbritrary error of $\pm$ 100 years to this point-estimate. However, finding TPQ dates that do not account for the likely distance between site and sea when the site was in use, limits, as with TPQ dates in general, the further inferential steps that can be taken. Furthermore, the more common practice of adding a constant error to a point-estimate does not take into account that the rate of shoreline displacement varies both spatially and temporally, and therefore introduces bias to the result on both the spatial and temporal scale, while also overestimating the precision of the method in most cases. This last point can be illustrated by finding the range between the upper and the lower limit of the four geological displacement curves for every meter from the marine limit and down to elevations that do not give an earliest date younger than 2500 BCE. This gives a mean TPQ range of `r round(tpqmean)` years with a standard deviation of `r round(tpqsd)` years across all four displacement curves, with this range varying significantly over time and between curves.

@@

When the frequency distribution of shoreline dates have been investigated in the past, this has typically involved finding point-estimates of dates by either using the mean of displacement curves, or by taking mean of the TPQ dates for each site (as illustrated in Figure \@ref(fig:shoredate)C). These point-estimates are then aggregated in bins of 200 or 500 years [e.g. @breivik2014; @solheim2018; @fossum2020; @mjærum2022; @tallavaara2020]. This approach is illustrated in Figure \@ref(fig:sumshore)B. The SSPD resulting from instead using the probabilistic method of shoreline dating, given in Figure \@ref(fig:sumshore)C, clearly demonstrates that accounting for the site-sea relationship gives a substantially different result compared to the more traditional approach, which is clearly indicated by the shifts in the positions of peaks and troughs in the distribution after c. 8000 BCE.

```{r sumshore, echo = FALSE, fig.cap = "A) The distribution of the shoreline dated sites (n = 921) by elevation. The binwidth is 4 m. B) The mean of \\textit{terminus post quem} dates for the sites, plotted with a binsize of 200 years, reflecting the common approach to summing shoreline dates. C) The result of instead summing the probability of the shoreline dates achievied with the method employed here, which accounts for the likely distance between sites and shoreline. The binwidth is 5 years, following the resolution used with the dating procedure.",  out.width = "500px", fig.align="center", regfloat = TRUE}

@@

The Monte Carlo simulations with the SSPD is here based on drawing a sample of calendar dates from the observed date range in the SSPD, equalling the number of shoreline dated sites, where the probability of drawing any 5 year interval is determined by the null model of choice. This is equivalent to the sampling method *calsample* from *rcarbon* [@crema2021]. Each sampled date is then assigned one of the pre-interpolated displacement curves, the probability of which is weighted by the density of observed sites within each 2 km interval. The calendar date is then 'uncalibrated' -- to follow the RSPD terminology -- to an elevation range from which a single elevation value is then drawn with uniform probability between the lower and upper limit of the displacement curve at that calendar year, using intervals of 5 cm. As shoreline dating is here done with a gamma distribution with the same parameters across all sites, there is no equivalent of the error term for ^14^C-dates that determine the shape of the Gaussian distribution associated with the ^14^C age. Consequently, the elevation value retrieved by 'uncalibrating' the shoreline date was then shoreline dated using the displacement curve for the relevant 2 km interval with the same gamma distribution for all samples. Having dated the number of samples equalling the number of shoreline dated sites, these were then summed, and the entire process repeated a total of 10000 times. The 97.5% highest and 2.5% lowest summed probability at each interval of 5 years across all simulations was then retrieved to create the 95% critical envelope, and the global p-value found in the same manner as for the RSPDs. This was done using re-purposed code from *rcarbon*.

## Continous piece-wise linear models
The procedures outlined above represent standard approaches when modelling population dynamics in archaeology. These can be helpful for determining whether the population behaviour has followed the trajectories of common theoretical population models, and is an especially useful exercise here as few assumptions concerning the behaviour and overall development of number of shorebound sites over time could be made. However, standard models frequently used to evaluate RSPDs are consistently being rejected in the literature, and deviations from these rejected null-models do not in themselves allow for any interpretations of the deviations, nor a subsequent direct analysis of the SPDs -- although this has frequently been the case. As a result, @timpson2021 have recently suggested an alternative framework for modelling past population dynamics. This involves fitting continuous piece-wise linear (CPL) models in addition to the standard models to the data, subjecting these to a procedure for model comparison, and finally testing the best model using the approach outlined above. The benefit of the CPL models is that these define discrete periods between hinge points that marks points in time in which the trajectory of relative population size changes direction, thus identifying key events and characteristics of the population dynamics between these. Following the evaluation of standard models outlined above, the data was therefore subsequently subjected to the approach outlined by @timpson2021.

@@

For each dating method, the relative performance of the exponential, logistic, uniform and CPL models was compared. This was done by finding the overall relative log-likelihoods of the models using the *loglik* function from *ADMUR*. For any given model, this involves finding the likelihood for each shoreline date against the null-model, and in the case of the ^14^C-dates, each site-phase. The overall model likelihood is the product of each of these individual likelihoods and is not, critically [@timpson2021], found by estimating the likelihood using the final SPD. From these likelihoods, the Bayesian (or Schwarz) information criterion (BIC) for each model was then found. This penalises the models for the number of parameters in use to avoid over-fitting.

@@

```{r spduniform, echo = FALSE, fig.cap = "Monte Carlo simulation for shoreline and radiocarbon dates. A) Summed probability of shoreline dated sites (n = 934) compared to a uniform null model. B) Same as A, but with a exponential model. C) Summed probability of calibrated radiocarbon dates (n =, bins = 678) compared to a uniform null model. D) Same as C, but with a exponential null model.", out.width = "500px", fig.align="center", regfloat = TRUE}
knitr::include_graphics(here("analysis/figures/mc.png"))
```
For the RSPD, the logistic model achieved a p-value of, and could therefore not be rejected. Based on the available data, this therefore represents the best explanation of the data. We can also add that accepting this model means that interpreting any deviations from the Monte Carlo enevelope is inappropriate, although this has sometimes been the case.

Comparing the the directions and deviations from the three standard models, Figure appears to indicate that some process of overall general increase or uniform distribution in the frequency of radiocarbon dates over time, contrasted to a general decrease in shoreline dated sites over time. However, with global p-values all below X,  all three models can be rejected as explaining the data they have been fit to, both for the SSPD and RSPD. Based on common demographic developments and previous suggestions, there is possibly some theoretical justification for expecting an exponential or logistic model to explain the shape of the RSPD, thus giving some further credence to interpreting deviations from this overall development as potentially meaningful signals of demographic developments. However, as introduced above, several authors have warned against over-interpreting such deviation. Furthermore, there are even less foundations on which to hypothesise what processes have dictated the development of site frequency over time, thus offering even further challenges to the interpretation of the SSPD and it's relationship with the null-models.

For the RSPD the rejected logistic model offers the best approximation of the data of the compared models. If taken as an indication of demographic developments this would indicate a first migration of people that is more substantial that what would be expected under the logistic null. Further deviations are then evident with the negative deviation and the poistive deviation c. These are, however, brief and can be left as little more than suggestions for possible meaningful demographic events.

In the following, some of these deviations will nonetheless be narratively linked to some major developments that have previously been proposed to characterise the Stone Age of south-eastern Norway, with the goal of suggesting some potential. It is, however, important to underscore that the *ad-hoc* coupling of the results achieved here with these developments should only be considered suggestions that can offer some potential avenues for further inquiry, for example using the approach outlined by @timpson2021.

@@

It is firmly established that Stone Age sites in the region date to the period from before the earliest ^14^C-dates, and at least from as far back as around 9300 BCE. While the SSPD starts from c. 9450 BCE, thus suggesting that the first date of human occupation in south-eastern Norway can be pushed back, this will require further corroboration from excavations and more meticulous evaluations of the evidence at hand. Furthermore, what has caused the lack of ^14^C-dates from this earliest period has been given many possible explanations, ranging from taphonomic loss, to diverse cultural factors, such as the potential use of rapidly deteriorating seal blubber for burning, which could also be related to a lack of reliable access to firewood. While taphonomic and ecological factors cannot be written off, we view it as likely that the initial jump in ^14^C-dates at the start of RSPD combined with the magnitude of the SSPD in the period preceding this, suggest that cultural factors is a central driver behind this pattern.

While all the considered models can be rejected as explaining the analysed data, they do nonetheless indicate that some process of decrease in shoreline dated site frequency occurs throughout the period, contrasted by a general stationary frequency of ^14^C-dated material. The most immediate explanation of this mismatch would be that the SSPD largely reflects variations in land-use and mobility patterns, while the RSPD mainly reflects population numbers. The modelled process of logistic decay markedly under-predicts the SSPD in the period X--X, and a deviation is also apparent in the period X--X, where the number of observed sites is significantly lower than what would be expected under the logistic model. While we are wary of over-interpreting these signals, they strike us as warranting more consideration than the other deviations that are barely outside the critical envelope, or are of a short duration. The deviations from after 4000 BCE should also be treated with further care, following both from questions concerning the reliability of shoreline dating after 4000 BCE [cf. @roalkvam2023].

The initial peak in the SSPD would be congruent with a process of rapid colonisation of the Norwegian coast, coupled with a high degree of mobility, leading to the overall high site-count in the earliest period. As evidenced by both genetic data and technological analyses of lithic inventories, this initial phase of human occupation appears to be followed by an influx of people from the . If this narrative is coupled to the data observed here, this process would appear to result in a relatively sudden drop in site-count as indicated in the SSPD, while the end of this drop corresponds to the first ^14^C-dates in the RSPD -- giving further credence to the suggestion that the appearance of material to be ^14^C-dated is related to cultural processes. In the period X--X following this, there is also a marked postive deviation from the RSPD under the exponential model, indicting a population boom, which, although barely breaking the critical envelope of the logitic model, is also evident in the SSPD.

@berg-hansen2022 have argued that there is an increase in sites in the first half of the MM in the northern part of the study area of this paper. However, their analysis is seemingly also based on the aggregation of point-based estimates of shoreline dates, which are aggregated within periods of variable length. Apart from the limitations with using point-based estimates, the use of disjoint time-intervals of variable duration could cause issues for comparing the relative counts across these units [@bevan2021]. The analysis undertaken here instead indicates a drop in the number of sites through this time-interval.   
While population numbers would likely impact the SSPD as well, the overall mismatch between the two proxies used here means that the magnitude of demographic influence on the SSPD is difficult to speculate about. However, the positive deviation in the SSPD from the logistic model at c. does correspond to a similar deviation of the RSPD from the exponential model. Matching negative deviations, on the other hand, can be observed around the middle of the 7th century BCE and in the period after 3500 BCE. While we are, as mentioned, cautions in giving these much weight in the SSPD alone, the correspondence of these deviations across the two proxies gives some reason to suggests that these might reflect meaningful demographic signals. This would also i

While it appears that the SSPD and RSPD are differentially impacted by population numbers, and, we suggest, the SSPD is more heavily influenced by land-use and mobility patterns, this suggestion and the nature of the discrepancy between the proxies clearly requires further investigation. Some lines of inquiry could be to also draw on lithic technology and tool-kit diversity as a function of population size and density [e.g. @riede2012; @solheim2020], and quantitative measures for mobility patterns that can potentially be inferred from the composition of lithic assemblages [e.g. @roalkvam2022]. Furthermore, environmental conditions, constraints and developments also represent a range of factors that can have implications for these dimensions, and will consequently be of future importance to unpack their relationship [e.g. @ordonez2022; @bergsvik2022; @jorgensen2020].

Finally, seeing as the method for shoreline dating implemented here was recently developed and has been subjected to limited evaluation and testing, there are still considerable uncertainties associated with its application [see @roalkvam2023]. While we still believe the approach represents an improvement over previous approaches, the results should therefore be treated with care. This also extends to the results of the RSPD, due to the fairly limited ^14^C sample. With these limitations in mind, the approach outlined by @timpson2021 as it offers a more conservative treatment of the data, reducing the amount of over-interpretation that might simply be the result of uncertainties to do with data or methods.

@@

>>>>

---
title: "Comparing summed probability distributions of shoreline and radiocarbon dates from the Mesolithic Skagerrak coast of Norway"
# author: 
# - 'Isak Roalkvam'
# - 'University of Oslo, Institute of Archaeology, Conservation and History'
  # - IsakRoalkvam:
  #     email: isak.roalkvam@iakh.uio.no
  #     institute: [iakh]
      # correspondence: true
#   - name: Steinar Solheim
#     email: steinar.solheim@khm.uio.no
#     institute: [mch]
#     correspondence: false
institute:
  - iakh: Institute of Archaeology, Conservation and History, University of Oslo
#   - mch: Museum of Cultural History, University of Oslo
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::pdf_document2:
      fig_caption: yes
      toc: false
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      # - --lua-filter=../templates/scholarly-metadata.lua
      # - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
header-includes:
  \usepackage{placeins}
# always_allow_html: true
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
# abstract: |
#   Text of abstract
# keywords: |
#   keyword 1; keyword 2; keyword 3
# highlights: |
#   These are the highlights. 
---

@@

# knitr::opts_knit$set(
#   eval.after = "fig.cap"
# )

@@

# Introduction
Population size is regarded as an important driver of cultural variation, and is of critical importance to our understanding of past human societies [e.g. @french2021; @shennan2000]. The frequency distribution of radiocarbon dates has been used extensively as one proxy for past relative population sizes [e.g. @crema2022; @french2021], including in Norwegian Stone Age archaeology [@solheim2018; @solheim2020a; @nielsen2019; @nielsen2021; @bergsvik2021; @jorgensen2020; @lundstrom2021; see also @persson2009]. The potentially immense value of insights into past population dynamics, combined with the ubiquity of radiocarbon dates and the relative ease with which these can now be processed within what has been termed a dates-as-data methodology [@rick1987] has undoubtedly contributed to the popularity of the approach. Several limitations and forms of criticism have, however, been directed at these procedures. Some of the objections are of a methodological nature, while others pertain to the underlying logic and the degree to which there is likely to be a direct connection between the frequency of ^14^C-dates and population dynamics [e.g. @carleton2021; @torfing2015; @ward2021]. What appears to be agreed upon by practitioners and critics alike is that radiocarbon dates are best analysed in this manner when compared and contrasted to other proxies for past population dynamics and other variables that might impact these [@french2016; @palmisano2017; @palmisano2021; @bergsvik2021]. This paper reports on and begins to unpack the relationship between two measures that have been linked to relative population size in the context of the Mesolithic Skagerrak coast in south-eastern Norway, namely summed probability distribution of calibrated radiocarbon dates and the summed probability distribution of shoreline dated sites (hereafter RSPD and SSPD, respectively).

@@

The final RSPD was subjected to the standard null-hypothesis testing approach through Monte Carlo simulation, as introduced above [see @shennan2013; @timpson2014], by fitting an exponential, logistic and uniform model to the observed RSPD. Following @timpson2021, the model parameters were identified by MLE search, using the differential evolution optimization algorithm *DEoptimR* [@brest2006]. The Monte Carlo simulations were then performed using the *SPDsimulationTest()* function from *ADMUR*. For each model, a series of individual calendar years are drawn from the model distribution, with replacement, the number of which equals the number of bins in the observed RSPD. These are then 'uncalibrated' to a single value on the ^14^C scale and a random error from among the observed errors is added to the date. These are then calibrated back to the calendar scale and finally summed. Here, this procedure was repeated 10000 times for each null model. The 2.5th and 97.5th quantile of the resulting summed probability for each year across all simulations are then retrieved to create the 95% critical envelope with which to compare the observed RSPD. The degree to which the observed RSPD deviates from the critical envelope is then used to calculate a global p-value, indicating whether or not the null can be rejected.

@@

Summing the probability of the shoreline dated sites and the model-fitting procedures followed the same structure as that for radiocarbon dates and was partly based on re-purposed programming code from *rcarbon* and *ADMUR*. However, idiosyncrasies in the dating method did necessitate some adjustments. To illustrate this, the procedure for shoreline dating a single site, as suggested in @roalkvam2023, is provided in Figure \@ref(fig:shoredate) and outlined below.

@@

With an estimated trajectory of shoreline displacement at the site to be dated, the elevation of the site can be drawn on to find a likely date for when it was in use, under the assumption that it was in use when located on or close to the shoreline. Figure \@ref(fig:shoredate)C indicates where the elevation of the example site intersects the interpolated displacement curve. As it can be reasonably be assumed that a site was not in use when located under water, this effectively represents a *terminus post quem* (TPQ) date. However, a fairly common approach to shoreline dating has been find where the elevation of the site intersects the mean of the upper and lower limit of the displacement curve, and then adding a somewhat arbitrary error of $\pm$ 100 years to this point-estimate. However, finding TPQ dates that do not account for the likely distance between site and sea when the site was in use limits the further inferential steps that can be taken. Furthermore, the more common practice of adding a constant error to a point-estimate does not take into account that the rate of shoreline displacement varies both spatially and temporally, and therefore introduces bias to the result on both the spatial and temporal scale, while also overestimating the precision of the method in most cases.

This last point can be illustrated by finding the range between the upper and the lower limit of the four geological displacement curves for every meter from the marine limit and down to elevations that do not give an earliest date younger than 2500 BCE. This gives a mean TPQ range of `r round(tpqmean)` years with a standard deviation of `r round(tpqsd)` years across all four displacement curves, with this range varying significantly over time and between curves.

@@

When the frequency distribution of shoreline dates have been investigated in the past, this has typically involved finding point-estimates of dates by either using the mean of displacement curves, or by taking mean of the TPQ dates for each site (as illustrated in Figure \@ref(fig:shoredate)C). These point-estimates are then aggregated in bins of 200 or 500 years [e.g. @breivik2014; @solheim2018; @fossum2020; @mjærum2022; @tallavaara2020]. This approach is illustrated in Figure \@ref(fig:sumshore)B. The SSPD resulting from instead using the probabilistic method of shoreline dating, given in Figure \@ref(fig:sumshore)C, demonstrates that accounting for the site-sea relationship gives a substantially different result compared to the more traditional approach, which is clearly indicated by the shifts in the positions of peaks and troughs in the distribution after c. 8000 BCE.

```{r sumshore, echo = FALSE, fig.cap = "A) The distribution of the shoreline dated sites (n = 921) by elevation. The binwidth is 4 m. B) The mean of \\textit{terminus post quem} dates for the sites, plotted with a binsize of 200 years, reflecting the common approach to summing shoreline dates. C) The result of instead summing the probability of the shoreline dates achieved with the method employed here, which accounts for the likely distance between sites and shoreline. The binwidth is 5 years, following the resolution used with the dating procedure.",  out.width = "500px", fig.align="center", regfloat = TRUE}

@@

The Monte Carlo simulations with the SSPD is here based on drawing a sample of calendar dates from the observed date range in the SSPD, equalling the number of shoreline dated sites, where the probability of drawing any 5 year interval is determined by the null model of choice. This is equivalent to the sampling method *calsample* from *rcarbon* [@crema2021]. Each sampled date is then assigned one of the pre-interpolated displacement curves, the probability of which is weighted by the density of observed sites within each 2 km interval. The calendar date is then 'uncalibrated' -- to follow the RSPD terminology -- to an elevation range from which a single elevation value is then drawn with uniform probability between the lower and upper limit of the displacement curve at that calendar year, using intervals of 5 cm. As shoreline dating is here done with a gamma distribution with the same parameters across all sites, there is no equivalent of the error term for ^14^C-dates that determine the shape of the Gaussian distribution associated with the ^14^C age. Consequently, the elevation value retrieved by 'uncalibrating' the shoreline date was then shoreline dated using the displacement curve for the relevant 2 km interval with the same gamma distribution for all samples. Having dated the number of samples equalling the number of shoreline dated sites, these were then summed, and the entire process repeated a total of 10000 times. The 97.5% highest and 2.5% lowest summed probability at each interval of 5 years across all simulations was then retrieved to create the 95% critical envelope, and the global p-value found in the same manner as for the RSPDs. This was done using re-purposed code from *ADMUR*.

## Continuous piece-wise linear models
The procedures outlined above represent standard approaches when modelling population dynamics in archaeology. These can be helpful for determining whether the population behaviour has followed the trajectories of common theoretical population models, and is an especially useful exercise here as few assumptions concerning the behaviour and overall development of number of shore-bound sites over time could be made. However, standard models frequently used to evaluate RSPDs are consistently being rejected in the literature, and deviations from these rejected null-models do not in themselves allow for any interpretations of the deviations, nor a subsequent direct analysis of the SPDs -- although this has frequently been the case. As a result, @timpson2021 have recently suggested an alternative framework for modelling past population dynamics. This involves fitting continuous piece-wise linear (CPL) models in addition to the standard models to the data, subjecting these to a procedure for model comparison, and finally testing the best model using the approach outlined above. The benefit of the CPL models is that these define discrete periods between hinge points that mark points in time in which the trajectory of relative population size changes direction, thus identifying key events and characteristics of the population dynamics between these. Following the evaluation of standard models outlined above, the data was therefore subsequently subjected to the approach outlined by @timpson2021.

@@

For each dating method, the relative performance of the exponential, logistic, uniform and CPL models was compared. This was done by finding the overall relative log-likelihoods of the models using the *loglik* function from *ADMUR*. For any given model, this involves finding the likelihood for each shoreline date against the null-model, and in the case of the ^14^C-dates, each site-phase. The overall model likelihood is the product of each of these individual likelihoods and is not, critically [@timpson2021, 3], found by estimating the likelihood using the final SPD. From these likelihoods, the Bayesian (or Schwarz) information criterion (BIC) was then found for each model. This penalises the models for the number of parameters in use, to avoid over-fitting, and can be used to compare the relative performance of the models.

@@

```{r spdmc, echo = FALSE, fig.cap = "Monte Carlo simulation for shoreline and radiocarbon dates. A) Summed probability of shoreline dated sites (n = 921) compared to an exponential null model. B) SSPD compared to a logistic null model. C) SSPD compared to a uniform null model. D) Summed probability of calibrated radiocarbon dates (n = 310, bins = 134) compared to an exponential null model. E) RSPD compared to a logistic null model. F) RSPD compared to a uniform null model.", out.width = "500px", fig.align="center", regfloat = TRUE}
knitr::include_graphics(here("analysis/figures/mc.png"))
```
For the RSPD, the logistic model achieved a p-value of, and could therefore not be rejected. Based on the available data, this therefore represents the best explanation of the data. For the SSPD all of the standard models could be rejected.

<!-- We can also add that accepting this model means that interpreting any deviations from the Monte Carlo enevelope is inappropriate, although this has sometimes been the case. -->

Comparing the the directions of the three standard models, Figure \@ref(fig:spd_mc) appears to indicate an overall uniform distribution in the frequency of radiocarbon dates following the initial jump just before 8000 BCE, contrasted to a general decrease in shoreline dated sites after the initial rapid increase from around 9500 BCE.

It is worth noting that there appears to be a corresponding peak in both proxies just after 8000 BCE and similarly a drop just before 6000 BCE. However, given that the logistic model cannot be rejected for the RSPD, it is inappropriate to interpret any deviations from the Monte Carlo envelope, and therefore no analytical weight can be placed on these deviations -- at least given the present data and modelling efforts.

@@

It is firmly established that Stone Age sites in the region date to the period from before the earliest ^14^C-dates, and at least from as far back as around 9300 BCE [e.g. @glorstad2016; @damlien2018b]. While the SSPD starts from c. 9500 BCE, thus suggesting that the first date of human occupation in south-eastern Norway can be pushed back, this will require further corroboration from excavations and more meticulous evaluations of the evidence at hand. Furthermore, what has caused the lack of ^14^C-dates from this earliest period has been given many possible explanations through the years, ranging from taphonomic loss to diverse cultural factors, such as the potential use of rapidly deteriorating seal blubber for burning, which could also be related to a lack of reliable access to firewood. While taphonomic and ecological factors cannot be written off, we view it as likely that the initial jump in ^14^C-dates at the start of RSPD combined with the magnitude of the SSPD in the period preceding this, suggest that cultural factors is a central driver behind this pattern.

The considered models indicate that after the initial jumps, some process of decrease in shoreline dated site frequency occurs throughout the period, contrasted by a general stationary frequency of ^14^C-dated material. The most immediate explanation of this mismatch would be that the SSPD largely reflects variations in land-use and mobility patterns, while the RSPD mainly reflects population numbers.

The development after 4000 BCE should be treated with further care, following from questions concerning the reliability of shoreline dating after 4000 BCE [cf. @roalkvam2023].

The initial peak in the SSPD would be congruent with a process of rapid colonisation of the Norwegian coast, coupled with a high degree of mobility, leading to the overall high site-count in the earliest period. As evidenced by both genetic data and technological analyses of lithic inventories, this initial phase of human occupation appears to be followed by an influx of people from the . If this narrative is coupled to the data observed here, this process would appear to result in a relatively sudden drop in site-count as indicated in the SSPD, while the end of this drop corresponds to the first ^14^C-dates in the RSPD -- giving further credence to the suggestion that the appearance of material to be ^14^C-dated is related to cultural processes.

@berg-hansen2022 have argued that there is an increase in sites in the first half of the MM in the northern part of the study area of this paper. However, their analysis is seemingly also based on the aggregation of point-based estimates of shoreline dates, which are aggregated within periods of variable length. Apart from the limitations with using point-based estimates for shoreline dates, the use of disjoint time-intervals of variable duration could cause issues for comparing the relative counts across these units [@bevan2021]. In contrast to their findings, the analysis undertaken here instead indicates a drop in the number of sites through this time-interval.

While population numbers would likely impact the SSPD as well, the overall mismatch between the two proxies used here means that the magnitude of demographic influence on the SSPD is not possible to ascertain.

<!-- However, the positive deviation in the SSPD from the logistic model at c. does correspond to a similar deviation of the RSPD from the exponential model. Matching negative deviations, on the other hand, can be observed around the middle of the 7th century BCE and in the period after 3500 BCE. While we are, as mentioned, cautions in giving these much weight in the SSPD alone, the correspondence of these deviations across the two proxies gives some reason to suggests that these might reflect meaningful demographic signals. This would also i -->

While it appears that the SSPD and RSPD are differentially impacted by population numbers, and, we suggest, the SSPD is more heavily influenced by land-use and mobility patterns, this suggestion and the nature of the discrepancy between the proxies clearly requires further investigation. Some lines of inquiry could be to also draw on lithic technology and tool-kit diversity as a function of population size and density [e.g. @collard2013a; @riede2014; @solheim2020], and quantitative measures for mobility patterns that can potentially be inferred from the composition of lithic assemblages [e.g. @roalkvam2022]. Furthermore, environmental conditions, constraints and developments also represent a range of factors that can have implications for these dimensions, and will consequently be of future importance to unpack their relationship [e.g. @ordonez2022; @jorgensen2020].

Finally, seeing as the method for shoreline dating implemented here was recently developed and has been subjected to limited evaluation and testing, there are still considerable uncertainties associated with its application. While we still believe the approach represents an improvement over previous approaches, the results should therefore be treated with care. This also extends to the results of the RSPD, due to the fairly limited ^14^C sample. With these limitations in mind, the approach outlined by @timpson2021 gives a more reasonable framework with which to evaluate the available data, as leads to a more conservative analysis and reduces the amount of over-interpretation that might follow from uncertainties to do with data or methods.

@@

# Ekstra figurer

Jeg fikk ADMUR til å funke med C14-dateringer. Figuren under viser at den logaritmiske modellen (blå kurve, samme som i Figur 5 over) ifølge BIC passer best av alternativene , og Monte Carlo-simuleringen over viser at den ikke kan avvises som forklarende for dataen. Eller med andre ord, dataen ville kunne forventes under den logaritmiske utviklingen. CPL modellering av strandlinjedateringene sliter jeg imidlertid mer med. Modellene opp til 4-CPL ser rimelige ut, men toppen på 5-CPL (sort kurve) som ligger litt før 5000 f.Kr. virker feil. I tillegg indikerer BIC-verdiene at denne passer best til dataen, som også virker rart gitt at den avviker såpass fra SSPDen.

```{r cpl_rcarbon, echo = FALSE, out.width = "500px", fig.align="center", regfloat = TRUE}
# fig.cap = "Model comparison, summed probability of radiocarbon dates.",
knitr::include_graphics(here("analysis/figures/cplrcarbon.png"))
```

```{r cpl_shore, echo = FALSE, out.width = "500px", fig.align="center", regfloat = TRUE}
# fig.cap = "Model comparison, summed probability of shoreline dates.",
knitr::include_graphics(here("analysis/figures/cplshore.png"))
```
